---
title: "A Bayesian Approach to Quarterback Accuracy"
subtitle: "STATS 115 Project"
author: 
  - Arnav Kanekar
  - Nicole Sonneville
  - Yvette Vargas
  - Allyson Lopez
date: "March 20, 2026"
format: revealjs
---

```{r}
library(tidyverse)
```

## Project Question

How can we estimate and compare the **true completion ability** of NFL quarterbacks by accounting for variability in **sample size** and **game conditions** using Bayesian statistics?

## Background

- Comparison of quarterbacks is a highly debated topic.
- Completion percentage is often used as a gauge of quarterback accuracy. 
  - $\text{Completion %} = \frac{\text{Number of Completions}}{\text{Number of Passes Attempted}}$
- But each quarterback faces unique and different scenarios and challenges throughout the season, so it is difficult to compare players directly.

Can we propose a method that allows NFL teams to get a better sense of the true ability of a quarterback?

## Data Selection

- Season long quarterback data
  - Official NFL data from 2015-2025 seasons
  - https://www.pro-football-reference.com

## Data Cleaning

- Cleaned data to only include data from players who have played enough 
  - kept passing data for only quarterbacks (main throwers of the ball in American football)
  - removed players who have started 3 games or less than 50 pass attempts
- Cleaned rows when a player appeared multiple times in one year
  - Due to team switches

## Dataset

- 452 rows, 15 columns, and 121 unique quarterbacks with varying occurrences over the decade of data
- Main variables: `Completions`, `Attempts`, `Comp_Percentage`
  - `Completions`: Number of passes completed in a season
  - `Attempts`: Number of passes attempted in a season
  - `Comp_Percentage`: Completion percentage, calculated by $\frac{\text{Number of Completions}}{\text{Number of Passes Attempted}}$
- Other variables include number of wins, number of touchdowns thrown, number of interceptions thrown, QB rating, number of sacks taken
  - These columns were not integral to our research question

## Dataset

```{r}
#| message: false
data <- read_csv("data/cleaned_data.csv")
data <- data %>%
  select(year, Player, Age, GS, Wins, Completions, Attempts, Comp_Percentage)

head(data, 10)
```

## Exploratory Data Analysis
```{r}
#2015–2024 
historical_qbs <- qb_data_clean %>%
  filter(year >= 2015, year <= 2024)


#2025 
qbs_2025 <- qb_data_clean %>%
  filter(year == 2025)

historical_qbs <- historical_qbs %>%
  mutate(comp_rate = Comp_Percentage / 100)
```


```{r}

ggplot(historical_qbs, aes(x = comp_rate)) +
  geom_histogram(bins = 21, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of QB Completion Rates (2015–2024)",
    x = "Completion Rate",
    y = "Number of QB Seasons"
  )
```



```{r}
historical_qbs %>%
  ggplot(aes(x = comp_rate)) +
  geom_density(fill = "lightblue", alpha = 0.5) +
  labs(
    title = "Density of QB Completion Rates",
    x = "Completion Rate",
    y = "Density"
  )
```

```{r}
historical_qbs %>%
  ggplot(aes(x = Attempts, y = comp_rate)) +
  geom_point(alpha = 0.5) +
  labs(
    title = "Completion Rate vs Pass Attempts",
    x = "Attempts",
    y = "Completion Rate"
  )
```

```{r}
historical_qbs %>%
  ggplot(aes(x = factor(year), y = comp_rate)) +
  geom_boxplot() +
  labs(
    title = "Completion Rate by Season",
    x = "Season",
    y = "Completion Rate"
  )
```

```{r}
qb_data_clean %>%
  ggplot(aes(x = Age)) +
  geom_histogram(bins = 15, fill = "skyblue", color = "white") +
  labs(
    title = "Distribution of QB Ages",
    x = "Age",
    y = "Number of QB Seasons"
  ) +
  theme_minimal()
```

```{r}
qb_data_clean %>%
  mutate(comp_rate = Completions / Attempts) %>%
  ggplot(aes(x = Age, y = comp_rate)) +
  geom_point(alpha = 0.5) +
  geom_smooth(method = "loess", se = FALSE, color = "red") +
  labs(
    title = "Completion Rate vs Age",
    x = "Age",
    y = "Completion Rate"
  ) +
  theme_minimal()
```


## Binomial Hierarchical Modeling

- We have correlation between seasons as most quarterbacks play more than one season.
- We will estimate random intercepts for each QB's true completion probability
  - shrink small-sample QBs toward the average


