## Modeling NFL Completion Rate Average

```{r}
#| message: false
library(tidyverse)
library(here)
library(rstanarm)
library(bayesplot)
library(broom.mixed)
library(loo)
```

```{r}
data <- read_csv(here("data", "cleaned_data.csv"))

# Should I model without 2025?
```

Simple Bayesian Binomial Model  
$y_i \sim \text{Binomial}(n_i, p)$  
$y_i =$ number of completions  
$n_i=$ number of pass attempts  
$p =$ probability of true completion  
```{r}
model <- stan_glm(
  cbind(Completions, Attempts - Completions) ~ 1,
  data = data,
  family = binomial(link = "logit"),
  prior = normal(0, 2.5),
  prior_intercept = normal(0, 5),
  chains = 4,
  iter = 2000,
  seed = 123,
  refresh = FALSE
)
summary(model)
```
The Rhat is close to 1 indicating good mixing.

$N_{eff} = 1600$ which is the effective number of independent posterior samples. This number of greater than 1000 which is good.

```{r}
mcmc_trace(model)
mcmc_dens_overlay(model)
mcmc_acf(model)
```

```{r}
tidy(model, effects = c("fixed", "aux"),
conf.int = TRUE, conf.level = 0.95)
```
With no predictors in the model, the median intercept estimates one overall completion probability for the entire dataset (the league average), which is 0.593 percent with a 95% credible interval between 0.583 and 0.602.

```{r}
data %>%
  summarise(actual_completion = mean(Completions/Attempts))
```
Based on the data, the actual completion rate from 2015-2025 is 63.8%

```{r}
pp_check(model, y_rep = 50)
```
Simulated data does not follow a similar distribution as the original data. The model does not seem to capture the variability and slight bimodality as the original distribution.

```{r}
loo(model)
```
`elpd_loo`, the expected log predictive density, is -2006.4. This number is valuable when comparing to future hierarchical model.
