---
title: "A Bayesian Approach to Quarterback Accuracy"
subtitle: "STATS 115 Project"
author: 
  - Arnav Kanekar
  - Nicole Sonneville
  - Yvette Vargas
  - Allyson Lopez
date: "March 20, 2026"
format: revealjs
---

```{r}
library(here)
library(tidyverse)
library(rstanarm)
library(bayesplot)
library(broom.mixed)
library(matrixStats)
library(tidybayes)
library(scales)
library(kableExtra)
```

## Project Question

How can we estimate and compare the **true completion ability** of NFL quarterbacks by accounting for variability in **sample size** and **game conditions** using Bayesian statistics?

## Background

**Passing** is almost synonymous with **winning** in the NFL.

-   **Pass**: A throw from one player to another.
-   **Completion**: A pass successfully caught by the receiver.

Quarterbacks **(QB)** are integral to a good passing team.

-   Comparison of NFL quarterbacks is a highly debated topic.

## [Why Raw Completion % Is Not Enough]{style="font-size: 0.75em;"}

-   Completion percentage is often a gauge of QB accuracy. $$\scriptsize \text{Completion %} = \frac{\text{Number of Completions}}{\text{Number of Pass Attempts}}$$
-   Direct comparison of QB completion rates is difficult since each QB faces unique scenarios and challenges.
    -   (ex: some QBs have better teams than others)

How can we use historical data to model a QB's "true" accuracy while accounting for situational variance?

## Data Selection

-   Season-long QB data
    -   Official NFL data from 11 seasons (2015-2025)
    -   <https://www.pro-football-reference.com>
    -   Contains many different passing related statistics

## Data Cleaning

-   Our clean data includes NFL players who:
    -   are QBs (main throwers of the ball in American football).
    -   have started more than 3 games.
    -   have more than 50 pass attempts.
-   We removed rows where players appeared multiple times in one year due to team switches.

## Dataset

-   452 rows, 15 columns, and 121 unique QBs with varying occurrences over the decade of data.
-   Main variables: `Completions`, `Attempts`, `Comp_Percentage`
    -   `Completions`: Number of passes completed in a season.
    -   `Attempts`: Number of passes attempted in a season.
    -   `Comp_Percentage`: Accuracy metric.

## Dataset Preview

```{r}
#| message: false
cleaned_data <- read_csv(here("data", "cleaned_data.csv"))
qb_data_clean <- cleaned_data %>%
  select(year, Player, Age, GS, Wins, Completions, Attempts, Comp_Percentage) %>%
  rename(Games = GS)

head(qb_data_clean, 10)
```

## Exploratory Data Analysis

```{r}
cleaned_data <- cleaned_data %>%
  mutate(Missed = Attempts - Completions) %>%
  select(Player, year, Comp_Percentage, Completions, Missed, Attempts, Age)

historical <- cleaned_data %>%
  filter(year < 2025)

data_2025 <- cleaned_data %>%
  filter(year == 2025)
```

```{r}
ggplot(historical, aes(x = Comp_Percentage / 100)) +
  geom_histogram(bins = 21, fill = "steelblue", color = "white") +
  labs(
    title = "Distribution of QB Completion Rates (2015–2024)",
    x = "Completion Rate",
    y = "Number of QB Seasons"
  ) +
  theme_minimal()
```

Most QBs cluster between **58–70%**, with the distribution roughly bell-shaped around \~64% and few outliers.

## Exploratory Data Analysis

```{r}
historical %>%
  ggplot(aes(x = Attempts, y = Comp_Percentage / 100, color = Age)) +
  geom_point(alpha = 1, size = 2) +
  scale_color_viridis_c() +
  labs(
    title = "Completion Rate vs Pass Attempts",
    x = "Number of Attempts",
    y = "Completion Rate"
  )
```

QBs with **more attempts stabilize** around the league mean, motivating **partial pooling** in our model.

## Exploratory Data Analysis

```{r}
historical %>%
  ggplot(aes(x = factor(year), y = Comp_Percentage / 100)) +
  geom_boxplot() +
  labs(
    title = "Completion Rate by Season",
    x = "Season",
    y = "Completion Rate"
  )
```

Median completion % has **gradually increased**, suggesting a league-wide shift toward better passing and justifying a **year-level random effect**.

## Binomial Hierarchical Modeling

QB seasons are correlated since QBs usually play several years.

A hierarchical model estimates completion percentage using **random intercepts** for each QB, shrinking players with few seasons toward the league average. This prevents overfitting on small samples.

## Building Our Prior

We used **2015–2024** to build a prior, then apply it to **2025** data. This lets us anchor our 2025 estimates to a decade of historical performance.

```{r}
#| echo: True
hist_model <- stan_glmer(
  cbind(Completions, Missed) ~ 1 + (1 | Player) + (1 | year),
  data = historical,
  family = binomial(link = "logit"),
  chains = 4,
  iter = 5000*2,
  seed = 123,
  refresh = FALSE
)
```

```{r}
#| echo: True
posterior <- as.data.frame(hist_model)
mu_median <- median(posterior$`(Intercept)`)
mu_sd   <- sd(posterior$`(Intercept)`)
```

## Building Our Posterior

We fit the same model on 2025 data only, plugging in `mu_median` and `mu_sd` from our prior.

```{r}
#| echo: True
model_2025 <- stan_glmer(
  cbind(Completions, Missed) ~ 1 + (1 | Player),
  data = data_2025,
  family = binomial(link = "logit"),
  prior_intercept = normal(location = mu_median, scale =  mu_sd),
  prior_aux = exponential(1),
  chains = 4,
  iter = 5000*2,
  seed = 123, 
  refresh = FALSE
)
```

## MCMC Trace Plot

```{r}
posterior_array <- as.array(model_2025)
```

```{r}
mcmc_trace(posterior_array, 
           pars = c("(Intercept)", "Sigma[Player:(Intercept),(Intercept)]"))
```

Chains are mixing well.

## MCMC Autocorrelation Plot

```{r}
mcmc_acf(posterior_array, 
         pars = c("(Intercept)", "Sigma[Player:(Intercept),(Intercept)]"))
```

Autocorrelation drops quickly to near zero.

## MCMC Posterior Densities

```{r}
mcmc_dens_overlay(posterior_array, 
           pars = c("(Intercept)", "Sigma[Player:(Intercept),(Intercept)]"))
```

All four chains have nearly identical density curves, confirming convergence to the same posterior.

## Predictive Posterior Check

```{r}
pp_check(model_2025, nreps = 100) +
  labs(title = "Posterior Predictive Check",
       subtitle = "Actual Completion % vs. 100 Simulated Distributions")
```

Simulated distributions generally follow the observed distribution.

## Player Random Intercepts

```{r}
# top 20 players
top_20_players <- data_2025 %>%
  group_by(Player) %>%
  slice(1) %>%
  add_epred_draws(model_2025) %>%
  median_qi(.epred) %>%
  ungroup() %>%
  slice_max(order_by = .epred, n = 20) %>%
  pull(Player)

# filter to just those players
data_top_20 <- data_2025 %>%
  filter(Player %in% top_20_players) %>%
  group_by(Player) %>%
  slice(1)

# posterior draws
player_previews_20 <- data_top_20 %>%
  add_epred_draws(model_2025)

ggplot(player_previews_20, aes(x = .epred, y = reorder(Player, .epred))) +
  # observed completion rate
  # 95% and 80% credible intervals
  stat_pointinterval(.width = c(.80, .95), color = "steelblue") +
  geom_point(
    data = data_top_20,
    aes(x = Comp_Percentage / 100, y = Player),
    color = "black", size = 2, stroke = 1.2
  ) +
  scale_x_continuous(
    labels = percent_format(accuracy = 1),
    breaks = seq(0, 1, by = 0.02)
  ) +
  labs(
    title = "Top 20 Quarterbacks: Predicted 2025 Completion Rates",
    subtitle = "95% and 80% Credible Intervals with Actual Rate in Black",
    x = "Expected Completion Percentage",
    y = NULL
  ) +
  theme_minimal() +
  theme(
    panel.grid.minor = element_blank(),
    axis.text.y = element_text(face = "bold"),
    axis.text.x = element_text(angle = 45, hjust = 1, vjust = 1)
  )
```

This plot displays the modeled completion rates and credible intervals for the top 20 QBs.

## [Top 10 Players: 2025 Data vs Our Model]{style="font-size: 0.75em;"}

```{r}
# 1. Extract posterior draws (rows = draws, cols = observations)
re_draws <- posterior_epred(model_2025, re.form = NULL)

# 2. Compute per-QB posterior median and 95% credible interval
qb_posterior <- data.frame(
  Player      = data_2025$Player,
  post_median = colMedians(re_draws),
  post_lower  = apply(re_draws, 2, quantile, 0.025),
  post_upper  = apply(re_draws, 2, quantile, 0.975)
)

# 3. Rank and select top 10
top_10_model <- qb_posterior %>%
  arrange(desc(post_median)) %>%
  slice(1:10)

top_10_data <- data_2025 %>%
  arrange(desc(Comp_Percentage)) %>%
  slice(1:10)

raw_tbl <- top_10_data %>%
  mutate(Rank = row_number()) %>%
  select(Rank, Player, Comp_Percentage) %>%
  rename(`Comp %` = Comp_Percentage)

model_tbl <- top_10_model %>%
  mutate(
    Rank       = row_number(),
    post_median = round(post_median * 100, 1)
  ) %>%
  select(Rank, Player, post_median, post_lower, post_upper) %>%
  rename(
    `Model %`  = post_median,
    `CI Lower` = post_lower,
    `CI Upper` = post_upper
  )

joined_tbl <- raw_tbl %>%
  full_join(model_tbl, by = "Player") %>%
  rename(Data_Rank = Rank.x, Model_Rank = Rank.y, CI_lo = `CI Lower`, CI_hi = `CI Upper`) %>%
  select(Player, Data_Rank, Model_Rank, `Comp %`, `Model %`, CI_lo, CI_hi)
```

```{r}
joined_tbl
```

We extract each QB's **posterior median completion probability**, then computing the median and 95% credible interval per player.

## Current Limitations

Computation Time

-   Random slopes for how completion percentage for each player changes over time
-   Add more predictors that reflect the game context or teammate performance rather than QB skill.

Statistical Model

-   Does not capture player improvement or age overtime.

## With Unlimited time and Resources

**More play-by-play data** including throw difficulty, defensive pressure, game context.

**Missing context** including defensive strength, receiver separation (distance), weather conditions, past/current injuries

**Stronger Hierarchical Model**

-   Add team, opponent, and situation effects
-   Time-varying component to model careers
-   Separate the true QB accuracy and external influences

## Impacts and Importance

-   Our analysis provides valuable insights into NFL passing trends and a new framework for evaluating QB accuracy.

    -   Our per season modeling provides insights of general NFL trends and how overall quarterback play has changed.

    -   Our per player modeling gives NFL teams a better understanding of a player’s true completion ability, which is valuable when comparing QBs. 
